### 2023-06-14 Manuscript/Draft1.tex
update manuscript

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption} \usepackage{array} \usepackage{caption} \usepackage{graphicx} \usepackage{siunitx} \usepackage{colortbl} \usepackage{multirow} \usepackage{hhline} \usepackage{calc} \usepackage{tabularx} \usepackage{tabulary} \usepackage{threeparttable} \usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

An Abstract

@@

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which sustained over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.). Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.

@@

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems with sampling design. Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the East coast.

@@

We construct spatial models for yellowtail flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question.

@@

\caption{Strata inhabited by each species in our population models. All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange. Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row. Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
\end{figure}

@@

The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation

@@

Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022). The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we rescaled the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.

@@

\caption{Visual representation of niche model for Cod.}\label{fig:hab-plot1}
\end{figure}

@@

\caption{Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Cod, and Haddock.}\label{fig:hab-plot}
\end{figure}

@@

Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume yellowtail flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include animated gif in final submission}

@@

In equation (1), \(Hab^2_{J,s}\) remains constant for each species for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week with temperature fluctuations. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation. Thus, stratified random samples in scenarios with a repeating temperature pattern will have constant survey catchability and availability over time, which may not be true for increasing temperature scenarios due to evolving spatial preferences.

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{cite}. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 \& 14 in the Spring and 37 \& 38 in the Fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.
An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

@@

We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values (\(Hab_{J.s}\)). When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b). After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A.

Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain. As seen right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time. By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption} \usepackage{array} \usepackage{caption} \usepackage{graphicx} \usepackage{siunitx} \usepackage{colortbl} \usepackage{multirow} \usepackage{hhline} \usepackage{calc} \usepackage{tabularx} \usepackage{tabulary} \usepackage{threeparttable} \usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

Many stock assessments in the United States use abundance estimates derived from stratified random bottom trawl data. To accurately represent true abundance, catches of a species must contain a low enough noise level to allow for a discernible pattern and the proportion of the population that is sampled should be consistent over time. These assumptions could be violated given enough noise in the sampling process and/or climate change causing a population to increase its abundance in strata that are not sampled. Using the R package MixFishSim, we have developed data-driven spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock in the western Atlantic Ocean to allow examination of these assumptions through simulation. Movement rates combine species-specific static habitat preferences with temperature tolerances. Habitat preferences were derived from niche models relating bottom trawl catches to environmental covariates. A repeating yearly temperature pattern produces repeating spatial biomass distributions in a given week, while a temperature gradient that increases on average over time results in spatial preferences that evolve throughout a given simulation. We created simulated spatial time series datasets for each species for several temperature scenarios and population trends. Using stratified random sampling on model output we were able to compare abundance estimates derived from the design-based stratified mean to those using a spatio-temporal model-based approach that allows inclusion of environmental covariates (VAST). Our focus was on the ability of contemporary indexing methods to track population trends under shifting spatial preferences.

@@

The continental shelf off the coast of the Northeast United States spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species, making commercial fisheries an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which supported over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.). Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.

@@

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process. Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.

@@

We construct spatial population models for Yellowtail Flounder (\emph{Limanda ferruginea}), Atlantic Cod (\emph{Gadus morhua}), and Haddock (\emph{Melanogrammus aeglefinus}) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock. Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. To address our research question we formulate the following inputs for the MFS package.

@@

\caption{Strata inhabited by each species in our population models. All strata on the eastern US contintental shelf are shown in the first image with Georges Bank highlighted in orange [@walsh2015long] **fix cite**. Stratum numbers used by the NEFSC bottom trawl survey are shown in the middle row. Strata that are excluded from certain stratified random sampling are shown in the bottom row in yellow.}\label{fig:strata-plot}
\end{figure}

@@

The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) in week \(wk+1\) using the formulation

@@

Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022). The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure seasonal influences to allow the niche model to instead infer static habitat preferences independent of temperature. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we rescaled the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.

@@

\caption{Visual representation of niche model for Atlantic Cod.}\label{fig:hab-plot1}
\end{figure}

@@

\caption{Static habitat preferences for each species in our population models. From left to right: Yellotwtail Flounder, Atlantic Cod, and Haddock.}\label{fig:hab-plot}
\end{figure}

@@

Each species is assumed to have normally distributed temperature preferences \(N(\mu,\sigma)\), with mean \(\mu\) and standard deviation \(\sigma\). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on spatial patterns while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include animated gif in final submission}

@@

In equation (1), \(Hab^2_{J,s}\) remains constant for each species for the duration of the simulation, while \(Tol_{c,s,wk}\) changes each week with temperature fluctuations. Using a temperature gradient that repeats every 52 weeks produces the same spatial preferences in a given week each year, resulting in relatively consistent spatial biomass patterns. Scenarios where the temperature increases over time creates spatial preferences that evolve as the water warms, producing spatial biomass patterns that shift in a given week over the duration of the simulation.

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. We consider increasing, decreasing, and constant population trends for Yellowtail for completeness. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 \& 14 in the spring and 37 \& 38 in the fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.

An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

@@

We follow the advice given in (Thorson 2019) and (Thorson, James T n.d.a) to build VAST estimates of biomass in our Georges Bank spatial population models using stratified random samples from our simulation output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. We considered the driving factors of movement in our population models as covariates- the dynamic temperature values and static habitat values (\(Hab_{J.s}\)). We ultimately decided to provide the most information to the model by including exact values for both temperature and habitat covariates to consider a best-case scenario.

Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b). After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.

Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \ref{fig:PopPct} depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain. By running simulations in \emph{MixFishSim} with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions. As seen in the right column of Figure \ref{fig:PopPct}, more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time.
